Light inextensible strings under tension in the Schwarzschild geometry

### 2020-10-22 string_physics_educator.tex
manuscript for The Physics Educator

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{units}
\usepackage{xcolor}
\modulolinenumbers[5]

\journal{The Physics Teacher}

\bibliographystyle{jss}

\begin{document}
\begin{frontmatter}
\title{Light inextensible strings under tension in the Schwarzschild geometry}

\begin{abstract}

Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.

% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\begin{keyword}
Black holes\sep light inextensible string\sep minimal-length path
\end{keyword}

\end{frontmatter}

\linenumbers
\section{Introduction}

@@

A light string's being under tension means that its path will be an
extremum of proper length.  In flat space, the calculus of variations
proves that the path is straight in the sense of zero curvature.  In
terrestrial gravity, standard $25\times
10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton sewing thread at its breaking
strain of about $40\unit{N}$ would have a sag of~$\sim 6\times
10^{-6}\meter$ over a $1\meter$ horizontal scale: barely perceptible.
How does such a quotidian observation translate into the vicinity of a
black hole?  Would the ``warping of space'' result in {\em visibly}
curved thread?

@@

@@

@@

@@

\bibliography{stringrefs.bib}

\clearpage

\section*{Figures}

\begin{figure}[h!] % Created by "flamm.R"
\centering
\includegraphics[width=\linewidth]{flamm_string.pdf}
\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
  path corresponding to a taut, light string}
\label{flamm}
\end{figure}

\begin{figure}[p] % Created by "closest_approach.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
\caption{Light inextensible strings under tension close to a black
  hole.  Here we see non-self intersecting strings arranged by
  increasing distance of closest approach to the event horizon,
  occurring tangentially at~$\phi=0$}
\label{closest_approach_non_self_intersecting}
\end{figure}

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
\caption{Light inextensible strings under tension close to a black
  hole.  Here self-intersecting strings are shown for $\phi\geqslant
  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

\begin{figure}[p] % Created by "light_closest_approach.R"
\centering
\includegraphics[width=\linewidth]{light_closest_approach.pdf}
\caption{Null geodesics in the Schwarzschild geometry, tangential
  at~$\phi=0$.  Note the differences between these curves and the taut
  strings shown elsewhere: unlike taut strings, null geodesices may
  cross the event horizon inwards, and are never tangential to the
  event horizon}
\label{light_closest_approach}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Light inextensible strings under tension close to a black
  hole.  Here strings passing through $(4M,0)$ at different angles are
  shown}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4M,0)$ at different angles,
  shown for $0\leqslant\phi\leqslant 2\pi$}
\label{light_r_equals_2}
\end{figure}

\end{document}

>>>>

% Authors may submit in single column format.
% Once your paper got accepted, the publisher will reformat it according to TPE house style (two column).
%\documentclass[wsdraft]{ws-tpe}

\documentclass{ws-tpe}
\usepackage{multicol}
\usepackage{units}
\usepackage[superscript]{cite}
\begin{document}

\markboth{Robin K. S. Hankin}{Light inextensible strings (thread) in the Schwarzschild metric}

\title{Light inextensible strings under tension in the Schwarzschild geometry}

\author{Robin K. S. Hankin}

\address{School of Engineering, Computing, and Mathematical Sciences, Auckland Iniversity of Technology, 55 Wellesley Street East\\
Auckland 1142, New Zealand
\email{hankin.robin@gmail.com}}

\maketitle

\begin{history}
\received{Day Month Year}
\revised{Day Month Year}
\end{history}

\begin{abstract}
Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.
\end{abstract}

\keywords{Black holes; Schwarzschild metric; light inextensible
  string; minimal-length path}

\section{Introduction}

@@

A light string's being under tension means that its path will be an
extremum of proper length.  In flat space, the calculus of variations
proves that the path is straight in the sense of zero curvature.  In
terrestrial gravity, standard $25\times
10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at its
breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
6\times 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely
perceptible.  How does such a quotidian observation translate into the
vicinity of a black hole?  Would the ``warping of space'' result in
{\em visibly} curved thread?

@@

\begin{figurehere} % Created by "flamm.R"
\centering
\includegraphics[width=\linewidth]{flamm_string.pdf}
\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
  path corresponding to a taut, light string}
\label{flamm}
\end{figurehere}

@@

\begin{figurehere} % Created by "closest_approach.R"
  \centerline{
    \includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
  }
  \caption{Light inextensible strings under tension close to a black
    hole.  Here we see non-self intersecting strings arranged by
    increasing distance of closest approach to the event horizon,
    occurring tangentially at~$\phi=0$}
  \label{closest_approach_non_self_intersecting}
\end{figurehere}

\begin{figurehere} % created by "closest_approach2.R"
  \centerline{
    \includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
  }
  \caption{Light inextensible strings under tension close to a black
    hole.  Here self-intersecting strings are shown for $\phi\geqslant
    0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
    o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
    about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
  \label{closest_approach_self_intersecting}
\end{figurehere}

\begin{figurehere} % Created by "light_closest_approach.R"
  \centerline{
    \includegraphics[width=\linewidth]{light_closest_approach.pdf}
  }
  \caption{Null geodesics in the Schwarzschild geometry, tangential
    at~$\phi=0$.  Note the differences between these curves and the taut
    strings shown elsewhere: unlike taut strings, null geodesices may
    cross the event horizon inwards, and are never tangential to the
    event horizon}
  \label{light_closest_approach}
\end{figurehere}

@@

\begin{figurehere}
  \centerline{
    \includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
  }
  \caption{Light inextensible strings under tension close to a black
    hole.  Here strings passing through $(4M,0)$ at different angles are
    shown}
  \label{strings_r_equals_2}
\end{figurehere}

\begin{figurehere}
  \centerline{
    \includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
  }
  \caption{Null geodesics passing through $(4M,0)$ at different angles,
    shown for $0\leqslant\phi\leqslant 2\pi$}
  \label{light_r_equals_2}
\end{figurehere}

@@

\bibliographystyle{unsrt}
\bibliography{stringrefs.bib}
\end{document}
